Distributional Regression Forests for Probabilistic Precipitation Forecasting in Complex Terrain

### 2018-10-17 pkg/disttree/inst/draft/draft.Rnw
some more revision based on reviews and feedback
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2328 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: fb83d2e9bc0fca7c239ad56c9b0d75e6f91707ab
some more revision based on reviews and feedback
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2331 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 1ae54badfd71bcdb3a85535a769dfad222b51798

\Abstract{
To obtain a probabilistic model for a dependent variable based on some set of
explanatory variables, a distributional approach is often adopted where the
parameters of the distribution are linked to regressors. In many classical
models this only captures the location of the distribution but
over the last decade there has been increasing interest in distributional
regression approaches modeling all parameters including location, scale, and
shape. Notably, so-called non-homogenous Gaussian regression (NGR) models both
mean and variance of a Gaussian response and is particularly
popular in weather forecasting. 
% More generally, the GAMLSS framework allows to establish generalized additive 
% models for location, scale, and shape with smooth linear or nonlinear effects.
Less specifically, generalized additive models for location, scale, and shape (GAMLSS) provide a framework where each distribution parameter is modeled separately capturing smooth linear or nonlinear effects.
%
However, when variable selection is required and/or there are
non-smooth dependencies or interactions (especially unknown or of high-order),
it is challenging to establish a good GAMLSS. A natural alternative in these
situations would be the application of regression trees or random forests but,
so far, no general distributional framework is available for these. Therefore,
a framework for distributional regression trees and forests is proposed that blends
regression trees and random forests with classical distributions from the GAMLSS
framework as well as their censored or truncated counterparts.
%
To illustrate these novel approaches in practice, they are employed to obtain
probabilistic precipitation forecasts at numerous sites in a mountainous
region (Tyrol, Austria) based on a large number of numerical weather
prediction quantities. It is shown that the novel distributional regression forests
automatically select variables and interactions, performing on par or often even
better than GAMLSS specified either through prior meteorological knowledge or a
computationally more demanding boosting approach.
}
\Keywords{parametric models, regression trees, random forests, recursive partitioning, probabilistic forecasting, GAMLSS}

@@

# add labels
text(x = -1.7, y = lh1, labels = "2009", col = pal[1], cex = 0.8)     # -0.8
text(x = -1.7, y = lh2, labels = "2010", col = pal[3], cex = 0.8)     # -0.8
text(x = -1.7, y = lh3, labels = "2011", col = pal[2], cex = 0.8)     # -0.8
text(x = -1.7, y = lh4, labels = "2012", col = pal[4], cex = 0.8)     # -0.8
@
\caption{\label{fig:axams-dist}\hlc{Total precipitation predictions by a 
distributional forest at station Axams for July 24 in 2009, 2010, 2011 and 
2012 learned on data from 1985--2008. Observations are non-negative and 
modeled by a Gaussian distribution left-censored at zero.
The observations are depicted by crosses and the predicted point mass
from the model by filled circles.}}
\end{figure}

@@

For obtaining probabilistic predictions from the tree for a (possibly new) set
of covariates $\bold{z} = (z_1, \ldots, z_m)$, the observation simply has to
be ``sent down'' the tree and the corresponding segment-specific MLE has to be
obtained. This can also be understood as a weighted MLE where the weights
select those observations from the learning sample that fall into the same
segment:
$$
w^{\text{tree}}_i(\bold{z}) = \sum_{b=1}^B \mathbf{1}((\bold{z}_i \in \mathcal{B}_b) \land (\bold{z} \in \mathcal{B}_b)),
$$
where $\mathbf{1}(\cdot)$ is the indicator function. The predicted distribution
for a given $\bold{z}$ is then fully specified by the estimated parameter
$\bm{\hat \theta}(\bold{z})$ where
$$
\bm{\hat \theta}(\bold{z}) = \argmax_{\bm{\theta} \in \bm{\Theta}} \sum_{i=1}^n w^{\text{tree}}_i(\bold{z}) \cdot \ell(\bm{\theta}; y_i).
$$
\hlc{Even though this parameter depends on the given $z$ it has to be noted that
it is the same for all other observations that are in the same terminal node of 
the tree as $z$ and therefore does not necessarily have to be recalculated for
each of them.}

@@

\hlc{The idea of random forests is to learn an ensemble of trees, each on a different
learning data obtained through resampling (bootstrap or subsampling). In each node 
only a random subset of the covariates $\bold{Z}$ is considered to reduce the 
correlation among the trees and to stabilizes the variance of the model.
For a simple regression random forest the mean of predictions over
all trees is considered. In that way changes in the location  of the response 
across the covariates are detected} (e.g. in Breiman and Cutler's random forests, 
\cite{Breiman:2001}).
\hlc{This idea is now taken one step further by embedding it in a distributional 
framework based on loglikelihood estimation. \emph{Distributional forests} 
employ an ensemble of $T$~\emph{distributional trees} which pick up changes in 
the ``direction'' of any distribution parameter by considering the full score 
vector for choosing splitting variables and split points.}

@@

Many statistical weather forecasting models leverage the strengths of modern
numerical ensemble prediction systems (EPSs; see \cite{Bauer+Thorpe+Brunet:2015}).
\hlc{EPSs not only predict the most likely future state of the atmosphere but provide
information about the uncertainty for a specific quantity and weather situation.
This is done by running the NWP model several times using slightly perturbed 
initial conditions and model specifications to account for uncertainties in both,
the initial state and the model and its parametrizations.
One frequently used method based on distributional regression models is the 
ensemble model output statistics (EMOS) approach first proposed by}
\cite{Gneiting+Raftery+Westveld:2005} \hlc{to produce high-quality forecasts for
specific quantities and sites.  
In case of precipitation forecasting EMOS typically uses the ensemble mean of 
``total precipitation''} (\emph{tp}) \hlc{forecasts as predictor for the location 
parameter $\mu$ and the corresponding ensemble standard deviation for the 
scale parameter $\sigma$ of the statistical model based on the assumption
of the response following a Gaussian distribution left-censored at zero.
This distributional approach of modeling both parameters allows to correct 
for possible errors of the NWP ensemble in both, the expectation but also the 
uncertainty of a specific forecast. In order to assure positivity of the 
scale parameter a log-link is applied. Hence, the two distribution parameters 
are modeled by the linear functions} 
$\mu = \beta_0 + \beta_1 \cdot \emph{mean}(\emph{tp})$ and  
$\log(\sigma) = \gamma_0 + \gamma_1 \cdot \emph{sd}(\emph{tp})$
with regression coefficients $\beta_0$, $\beta_1$, $\gamma_0$ and $\gamma_1$,  
\hlc{hereafter referred to as a basic EMOS model} \citep{Gebetsberger+Messner+Mayr:2017}.
%$\mu = \beta_0 + \beta_1 \cdot \mbox{mean}(\mbox{tp}^{\frac{1}{p}})$ and  
%$\log(\sigma) = \gamma_0 + \gamma_1 \cdot \mbox{sd}(\mbox{tp}^{\frac{1}{p}})$
%with power transformation parameter $\frac{1}{p}$ and 
%regression coefficients $\beta_0$, $\beta_1$, $\gamma_0$ and $\gamma_1$,  
%\hlc{hereafter referred to as a basic EMOS model \citep{Gebetsberger+Messner+Mayr:2017}.}

\hlc{\textit{FIXME: notation with or without power transformation parameter?}}

While this approach alone is already highly effective in the plains,
it typically does not perform as well in complex terrain due to unresolved
effects in the NWP system \citep{Bauer+Thorpe+Brunet:2015}. 
For example, in the Tyrolean Alps -- considered
in the following case study -- the NWP grid cells of $50 \times 50$
km$^2$ are too coarse to capture single mountains, narrow valleys, etc. 
Therefore, it is often possible to substantially improve the predictive 
performance of the EMOS by including additional predictor variables, either 
from local meteorological observations or an NWP model. Unfortunately, it is typically unknown 
which variables are relevant for improving the predictions. Simply including
all available variables may be computationally burdensome and
can lead to overfitting but, on the other hand, excluding too many variables
may result in a loss of valuable information. Therefore, selecting
the relevant variables and interactions among all possible covariates is 
crucial for improving the statistical forecasting model.

@@

\item \emph{EMOS:} The basic ensemble model output statistics
  models use the ensemble mean of total precipitation as regressor in the location
  submodel and the corresponding ensemble standard deviation in the scale submodel.
  The parameters are estimated by maximum likelihood, using an identity
  link for the location part and a log link for the scale part
  \citep[following the advice of][]{Gebetsberger+Messner+Mayr:2017}.

@@

As stated by \cite{Breiman:2001} and \cite{Hastie+Tibshirani+Friedman:2001} 
\hlc{random forests do not overfit, not even for an increasing number of 
trees} (\code{ntree}). \hlc{Hence, it is recommended to build forests consisting 
of a large number of trees as this generally leads to an increase 
in accuracy. However, it also comes along with increasing computational costs. 
Therefore, ``a good choice results from a trade-off between computational 
complexity and accuracy''} \citep{Biau+Scornet:2016}.
\hlc{Following this advice we chose to build forests consisting of 100 trees.}

@@

The fitted distributional forest for July 24 and
observation station Axams (including Figure~\ref{fig:axams-dist})
is reproducible using 
\code{demo("RainAxams", package = "disttree")}.
This also includes fitting the other zero-censored
Gaussian models considered in this paper and generating the corresponding 
QQ~plots (Figure~\ref{fig:axams-qq}).
Full replication of all results can be obtained with
\code{demo("RainTyrol", package = "disttree")}
requiring the companion \textsf{R} package \textbf{RainTyrol} (version~\Sexpr{packageVersion("RainTyrol")}),
also available within the \textsf{R}-Forge~project.

>>>>

\Abstract{
To obtain a probabilistic model for a dependent variable based on some set of
explanatory variables, a distributional approach is often adopted where the
parameters of the distribution are linked to regressors. In many classical
models this only captures the location of the distribution but
over the last decade there has been increasing interest in distributional
regression approaches modeling all parameters including location, scale, and
shape. Notably, so-called non-homogenous Gaussian regression (NGR) models both
mean and variance of a Gaussian response and is particularly
popular in weather forecasting. 
% More generally, the GAMLSS framework allows to establish generalized additive 
% models for location, scale, and shape with smooth linear or nonlinear effects.
\hlc{Less specifically, generalized additive models for location, scale, and shape (GAMLSS) provide a framework where each distribution parameter is modeled separately capturing smooth linear or nonlinear effects.}
%
However, when variable selection is required and/or there are
non-smooth dependencies or interactions (especially unknown or of high-order),
it is challenging to establish a good GAMLSS. A natural alternative in these
situations would be the application of regression trees or random forests but,
so far, no general distributional framework is available for these. Therefore,
a framework for distributional regression trees and forests is proposed that blends
regression trees and random forests with classical distributions from the GAMLSS
framework as well as their censored or truncated counterparts.
%
To illustrate these novel approaches in practice, they are employed to obtain
probabilistic precipitation forecasts at numerous sites in a mountainous
region (Tyrol, Austria) based on a large number of numerical weather
prediction quantities. It is shown that the novel distributional regression forests
automatically select variables and interactions, performing on par or often even
better than GAMLSS specified either through prior meteorological knowledge or a
computationally more demanding boosting approach.
}
\Keywords{parametric models, regression trees, random forests, recursive partitioning, probabilistic forecasting, GAMLSS}

@@

# add labels
text(x = -1.7, y = lh1, labels = "2009", col = pal[1], cex = 0.8)     # -0.8
text(x = -1.7, y = lh2, labels = "2010", col = pal[3], cex = 0.8)     # -0.8
text(x = -1.7, y = lh3, labels = "2011", col = pal[2], cex = 0.8)     # -0.8
text(x = -1.7, y = lh4, labels = "2012", col = pal[4], cex = 0.8)     # -0.8
@
\caption{\label{fig:axams-dist}\hlc{Total precipitation predictions by a 
distributional forest at station Axams for July~24 in 2009, 2010, 2011 and 
2012 learned on data from 1985--2008. Observations are non-negative and 
modeled by a Gaussian distribution left-censored at zero.
The observations are depicted by crosses and the predicted point mass
from the model by filled circles.}}
\end{figure}

@@

For obtaining probabilistic predictions from the tree for a (possibly new) set
of covariates $\bold{z} = (z_1, \ldots, z_m)$, the observation simply has to
be ``sent down'' the tree and the corresponding segment-specific MLE has to be
obtained. This can also be understood as a weighted MLE where the weights
select those observations from the learning sample that fall into the same
segment:
$$
w^{\text{tree}}_i(\bold{z}) = \sum_{b=1}^B \mathbf{1}((\bold{z}_i \in \mathcal{B}_b) \land (\bold{z} \in \mathcal{B}_b)),
$$
where $\mathbf{1}(\cdot)$ is the indicator function. The predicted distribution
for a given $\bold{z}$ is then fully specified by the estimated parameter
$\bm{\hat \theta}(\bold{z})$ where
$$
\bm{\hat \theta}(\bold{z}) = \argmax_{\bm{\theta} \in \bm{\Theta}} \sum_{i=1}^n w^{\text{tree}}_i(\bold{z}) \cdot \ell(\bm{\theta}; y_i).
$$
\hlc{Even though this parameter depends on the given $\bold{z}$ it has to be noted that
it is the same for all other observations that are in the same terminal node of 
the tree as $\bold{z}$ and therefore does not necessarily have to be recalculated for
each of them.}

@@

\hlc{The idea of random forests is to learn an ensemble of trees, each on a different
learning data obtained through resampling (bootstrap or subsampling). In each node 
only a random subset of the covariates $\bold{Z}$ is considered to reduce the 
correlation among the trees and to stabilize the variance of the model.
For a simple regression random forest the mean of predictions over
all trees is considered. In that way changes in the location  of the response 
across the covariates are detected} (e.g. in Breiman and Cutler's random forests, 
\cite{Breiman:2001}).
\hlc{This idea is now taken one step further by embedding it in a distributional 
framework based on loglikelihood estimation. \emph{Distributional forests} 
employ an ensemble of $T$~\emph{distributional trees} which pick up changes in 
the ``direction'' of any distribution parameter by considering the full score 
vector for choosing splitting variables and split points.}

@@

Many statistical weather forecasting models leverage the strengths of modern
numerical ensemble prediction systems (EPSs; see \cite{Bauer+Thorpe+Brunet:2015}).
\hlc{EPSs not only predict the most likely future state of the atmosphere but provide
information about the uncertainty for a specific quantity and weather situation.
This is done by running the NWP model several times using slightly perturbed 
initial conditions and model specifications to account for uncertainties in both,
the initial state and the model and its parametrizations.
One frequently used method based on distributional regression models is the 
ensemble model output statistics (EMOS) approach first proposed by}
\cite{Gneiting+Raftery+Westveld:2005} \hlc{to produce high-quality forecasts for
specific quantities and sites.  
In case of precipitation forecasting EMOS typically uses the ensemble mean of 
``total precipitation''} (\emph{tp}) \hlc{forecasts as predictor for the location 
parameter $\mu$ and the corresponding ensemble standard deviation for the 
scale parameter $\sigma$ of the statistical model based on the assumption
of the response following a Gaussian distribution left-censored at zero.
This distributional approach of modeling both parameters allows to correct 
for possible errors of the NWP ensemble in both, the expectation but also the 
uncertainty of a specific forecast. In order to assure positivity of the 
scale parameter a log-link is applied. Hence, the two distribution parameters 
are modeled by the linear functions} 
$\mu = \beta_0 + \beta_1 \cdot \emph{mean}(\emph{tp})$ and  
$\log(\sigma) = \gamma_0 + \gamma_1 \cdot \emph{sd}(\emph{tp})$
with regression coefficients $\beta_0$, $\beta_1$, $\gamma_0$ and $\gamma_1$,  
\hlc{hereafter referred to as a basic EMOS model} \citep{Gebetsberger+Messner+Mayr:2017}.

% FIXME: notation with or without power transformation parameter?

% with:
%$\mu = \beta_0 + \beta_1 \cdot \mbox{mean}(\mbox{tp}^{\frac{1}{p}})$ and  
%$\log(\sigma) = \gamma_0 + \gamma_1 \cdot \mbox{sd}(\mbox{tp}^{\frac{1}{p}})$
%with power transformation parameter $\frac{1}{p}$ and 
%regression coefficients $\beta_0$, $\beta_1$, $\gamma_0$ and $\gamma_1$,  
%\hlc{hereafter referred to as a basic EMOS model \citep{Gebetsberger+Messner+Mayr:2017}.}

While this approach alone is already highly effective in the plains,
it typically does not perform as well in complex terrain due to unresolved
effects in the \hlc{NWP system} \citep{Bauer+Thorpe+Brunet:2015}. 
For example, in the Tyrolean Alps -- considered
in the following case study -- the NWP grid cells of $50 \times 50$
km$^2$ are too coarse to capture single mountains, narrow valleys, etc. 
Therefore, it is often possible to substantially improve the predictive 
performance of the EMOS by including additional predictor variables, either 
from local meteorological observations or an NWP model. Unfortunately, it is 
typically unknown which variables are relevant for improving the predictions. 
Simply including all available variables may be computationally burdensome and
can lead to overfitting but, on the other hand, excluding too many variables
may result in a loss of valuable information. Therefore, selecting
the relevant variables and interactions among all possible covariates is 
crucial for improving the statistical forecasting model.

@@

\item \emph{EMOS:} \hcl{The basic} ensemble model output statistics
  models use the ensemble mean of total precipitation as regressor in the location
  submodel and the corresponding ensemble standard deviation in the scale submodel.
  The parameters are estimated by maximum likelihood, using an identity
  link for the location part and a log link for the scale part
  \citep[following the advice of][]{Gebetsberger+Messner+Mayr:2017}.

@@

As stated by \cite{Breiman:2001} and \cite{Hastie+Tibshirani+Friedman:2001} 
\hlc{random forests do not overfit, not even for an increasing number of 
trees} (\code{ntree}). \hlc{Hence, it is recommended to build forests consisting 
of a large number of trees as this generally leads to an increase 
in accuracy. However, it also comes along with increasing computational costs. 
Therefore, ``a good choice results from a trade-off between computational 
complexity and accuracy''} (\cite{Biau+Scornet:2016}, p.~205).
\hlc{Following this advice we chose to build forests consisting of 100 trees.}

@@

The fitted distributional forest for July 24 and
observation station Axams (including Figure~\ref{fig:axams-dist})
is reproducible using 
\code{demo("RainAxams", package = "disttree")}.
This also includes fitting the other zero-censored
Gaussian models considered in this paper and generating the corresponding 
QQ~plots (Figure~\ref{fig:axams-qq}) and PIT histograms 
(Supplement 2: Stationwise Evaluation).
Full replication of all results can be obtained with
\code{demo("RainTyrol", package = "disttree")}
requiring the companion \textsf{R} package \textbf{RainTyrol} (version~\Sexpr{packageVersion("RainTyrol")}),
also available within the \textsf{R}-Forge~project.

### 2018-10-17 pkg/disttree/inst/draft/rejoinder.Rnw
some more revision based on reviews and feedback
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2331 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 1ae54badfd71bcdb3a85535a769dfad222b51798

\newenvironment{reto}{
    \begin{color}{uibkorange}
    \textbf{Reto:~}
        \itshape
}{
    \end{color}
}
\usepackage{amsmath}

@@

The skewness results from the characteristics of precipitation sums. 
Most of the time no or only moderate precipitation amounts are forecasted 
and observed. In some situations (e.g., days with thunderstorms, cold and 
warm fronts, \dots) large amounts of precipitation can be observed at 
specific sites. This is often also depict by the ensemble prediction 
system where a subset of all members predicts rather large amounts 
(compared to the overall ensemble mean or median). This yields strongly 
skewed ensemble forecasts which has to be addressed in some sort.\\
As you mention one way would be to directly apply a
distributional family with an additional parameter to capture
skewness. Another frequently used approach is to transform the 
data to remove large parts of the skewness. The latter has the 
advantage that the model assumptions get simpler as no estimate 
for the skewness parameter is required. 
Moreover, applying a transformation together with a Gaussian distribution
censored at zero allows for a better comparison to the other models as 
they can all be built under the same settings.
In this study we are thus following Stauffer et al. (2017) 
using the power transformation which turned out to be optimal 
for a similar application for the very same study area.

@@

This is the trade-off between interpretability and predictive 
skill. Regression-type models might be easier to interpret, 
however, if one is mainly interested in predictive skill and
a good performing fully automated weather forecasting tool the 
interpretability is not implicitly needed in some cases 
(machine learning performance vs. interpretability).
We adapted our manuscript as we definitely want to emphasize that 
in our application the focus is mainly on prediction but not on 
interpretation. We hope we could point this out clearly.

@@

\item[] Hastie, T., Tibshirani, R. and Friedman, J. (2001).
The Elements of Statistical Learning, Second Edition,
\textit{Springer New York Inc.}.

>>>>

\newenvironment{reto}{
    \begin{color}{uibkorange}
    \textbf{Reto:~}
        \itshape
}{
    \end{color}
}

\definecolor{fixmecol}{rgb}{1,1,0.7}
\newcommand{\fixme}[1]{{\sethlcolor{fixmecol}\hl{#1}}}

\usepackage{amsmath}

@@

The skewness results from the characteristics of precipitation sums. 
Most of the time no or only moderate precipitation amounts are forecasted 
and observed. In some situations (e.g., days with thunderstorms, cold and 
warm fronts, \dots) large amounts of precipitation can be observed at 
specific sites. This is often also depict by the ensemble prediction 
system where a subset of all members predicts rather large amounts 
(compared to the overall ensemble mean or median). This yields strongly 
skewed ensemble forecasts which has to be addressed in some sort.\\
\fixme{\textit{FIXME: Should this explanation why there is skewness be included?}}

As you mention one way how to deal with skewness would be to 
apply a distributional family with an additional parameter to capture
skewness. However, this can be very complex and thus lead to less well 
fitting estimators of the location and scale parameter.
Moreover, the focus of this paper is on the features of the presented 
distributional forest regarding variable selection and the capability 
to also model non-linear effects rather than optimizing the model 
specification for this particular application.
Transforming the data to remove large parts of the skewness is
a frequently used approach. Therefore, employing such a transformation 
allows for a better way of comparing our results to those presented in 
literature. In this study we are thus following Stauffer et al. (2017) 
using the power transformation which turned out to be optimal 
for a similar application for the very same study area.

@@

We completely agree with you as that other models such as
GAMLSS are clearly advantageous regarding interpretability. 
These models provide a very powerful framework that works
well in a wide range of applications and research fields, 
including meteorology. Therefore, our intention is not to present 
distributional forests as a replacement of these commonly used methods 
but rather as a well working alternative. Especially if one is mainly 
interested in predictive skill and a good performing fully automated 
weather forecasting tool the interpretability is not implicitly needed 
in some cases (machine learning performance vs. interpretability).
We adapted our manuscript as we definitely want to emphasize that in our 
application the focus is mainly on prediction but not on interpretation. 
We hope we could point this out clearly. \\
\fixme{\textit{FIXME: Where should this be pointed out in the manuscript? Discussion?}}

@@

\item[] Hastie, T., Tibshirani, R. and Friedman, J. (2001).
The Elements of Statistical Learning, %Second Edition,
\textit{Springer New York Inc.}.
